Can Auxiliary Indicators Improve COVID-19 Forecasting and Hotspot Prediction?

### 2021-06-18 forecast/paper/paper.tex
Edits to up, down, flat

\begin{figure}[t]
\includegraphics[width=\columnwidth]{fig/ny-trajectory-1.pdf}
\caption{Forecast from the autoregressive model for the New York City HRR made
  on October 15 (vertical line).  The fan displays 50\%, 80\% and 95\% intervals
  while the orange curve shows the median forecast. The connected black points
  represent ``finalized'' data, as reported in May 2021.}
\label{fig:trajectory}
\end{figure}

@@

In addition to this pure autoregressive model, we also consider five
logistic models of the form:  
\begin{multline}
\label{eq:hotspot_ar_x}
\logit \big( \P(Z_{\ell,t+a} = 1 \,|\, Y_{\ell,s}, X_{\ell,s}, s \leq t) \big)
= \\ \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y^\Delta_{\ell,t-7j} +  
\sum_{j=0}^2 \gamma^{a,\tau}_j X^\Delta_{\ell,t-7j},
\end{multline}
where we use \smash{$X^\Delta_{\ell,t} = (X_{\ell,t} - X_{\ell,
    t-7})/(X_{\ell,t-7})$}, and again $X_{\ell,t}$ stands for any of the five  
auxiliary indicators at location $\ell$ and time $t$.  We fit all models
\eqref{eq:hotspot_ar}, \eqref{eq:hotspot_ar_x} using logistic regression, 
pooling all 306 HRRs and using a 21-day trailing window for the training data.

@@

WIS is typically defined for quantile-based forecasts where the quantile levels
are symmetric around 0.5.  This is the case for our choice in
\eqref{eq:quantile_levels}.  Let $F$ be a forecaster comprised of predicted
quantiles $q_\tau$ parametrized by a quantile level $\tau$.  In the case of
symmetric quantile levels, this is equivalent to a collection of central
prediction intervals $(\ell_\alpha, u_\alpha)$, parametrized by an exclusion
probability $\alpha$. The WIS of the forecaster $F$, evaluated at the target
variable $Y$, is defined by:
\begin{equation}
\label{eq:wis_intervals}
\WIS(F,Y) = \sum_\alpha \Big\{\alpha(u_\alpha - \ell_\alpha) + 2 \cdot
\dist(Y, [\ell_\alpha, u_\alpha])\Big\},  
\end{equation}
where $\dist(a, S)$ is the distance between a point $a$ and set $S$ (the
smallest distance between $a$ and an element of $S$).  Note that, corresponding 
to \eqref{eq:quantile_levels}, the exclusion probabilities are $\alpha \in
\{0.05, 0.2, 0.5, 1\}$, resulting in 4 terms in the above sum.  By 
straightforward algebra, it is not hard to see WIS has an alternative 
representation in terms of the predicted quantiles themselves:
\begin{equation}
\label{eq:wis_quantiles}
\WIS(F,Y) = 2 \sum_\tau \phi_\tau(Y - q_\tau), 
\end{equation}
where $\phi_\tau(x) = \tau |x|$ for $x \geq 0$ and $\phi_\tau(x) = (1-\tau)
|x|$ for $x<0$, which is often called the ``tilted absolute'' loss.  While
\eqref{eq:wis_quantiles} is more general (it can accomodate asymmetric quantile
levels), the first definition \eqref{eq:wis_intervals} is usually preferred in
presentation, because the score nicely decouples into a ``sharpness'' component
(first term in each summand) and ``under/overprediction'' component (second term
in each summand).  The second form given in \eqref{eq:wis_quantiles} is
especially noteworthy in our current study as it reveals WIS is the same as the
quantile regression loss that we use to train our forecasting models (i.e., our
models are fit to optimize WIS averaged over the training set).

@@

We begin with a summary of the high-level conclusions.  Here, and in the
following subsections, we will use ``AR'' to refer to the pure autoregressive
model both in forecasting \eqref{eq:forecast_ar} and in hotspot prediction
\eqref{eq:hotspot_ar} (where the reference to the task should be clear from the
context). We will also use the name of an auxiliary indicator---``CHNG-CLI'', 
``CHNG-CLI'', ``CHNG-COVID'', ``CTIS-CLI-in-community'', ``DV-CLI'', or
``Google-AA''---interchangeably with the model in forecasting
\eqref{eq:forecast_ar_x} or hotspot prediction \eqref{eq:hotspot_ar_x} that uses
this particular indicator as a feature (the meaning should be clear from the
context).  So, e.g., the CHNG-CLI model in forecasting is the one in
\eqref{eq:forecast_ar_x} that sets $X_{\ell,t}$ to be the value of the CHNG-CLI
indicator at location $\ell$ and time $t$, fits the coefficients based on the
training data, and makes its predictions by setting $X_{\ell,t}$ to the
appropriate CHNG-CLI values.  Finally, we use the term ``indicator model'' to 
refer to any one of the ten models of the form \eqref{eq:forecast_ar_x},
\eqref{eq:hotspot_ar_x} (five from each of the forecasting and hotspot
prediction tasks).

@@

Figure~\ref{fig:forecast} displays the results for forecasting, stratified by
ahead value and averaged over all HRRs and forecast dates (June 9--December 31,
2020).  Shown is the average WIS for each forecast model divided by the averaged
WIS from a baseline model, which is basically a flat-line forecaster (its
median forecast for $Y_{\ell,t+a}$ is always $Y_{\ell,t}$, with predicted
quantiles defined around this flat line based on historical variation). This is
the same baseline model as that used in the COVID-19 Forecast Hub.  Here, we  
divide by the average WIS of the baseline model in order to put the y-axis on an 
interpretable scale.  We can see that all curves are well below 1, which means
(smaller WIS is better) that all of the models, including AR, outperform the
baseline on average over the forecast period (though this advantage narrows as
the ahead value grows).

@@

\begin{figure}[t]
  \includegraphics[width=\columnwidth]{fig/fcast-1.pdf}
  \caption{Average WIS for each forecast model divided by the averaged WIS of
    the flat-line forecaster.}
  \label{fig:forecast}
\end{figure}

Figure~\ref{fig:hotspot} displays the results for hotspot prediction, again
stratified by ahead value and averaged over all HRRs and prediction dates (June 
16--December 31, 2020).  We can see many similarities to the forecasting
results (recall larger AUC is better): CHNG-COVID and DV-CLI offer the biggest
improvement over AR, all models including AR degrade in performance towards the
baseline (this is now just a classifier based on random guessing, which achieves
an AUC of 0.5) as the ahead values grow.  However, one clear difference is that
the CHNG-CLI model performs quite well in hotspot prediction, close to the
best-performing indicator models for many of the ahead values.

[TODO: should we
insert one line explaining this---you're somewhat protected form volatiliy in a
logistic model thanks to 
the sigmoid function used to define the probabilities?]

@@

To test this implicit regularization hypothesis, we reran all of the prediction
experiments but with $X_{\ell,t}$ in each indicator model \eqref{eq:forecast_ar_x},
\eqref{eq:hotspot_ar_x} replaced with suitable random noise (bootstrap
samples form an indicator's historical distribution).  The results, shown and
explained more precisely in the supplement, are vastly different (worse) than
the original set of results.  In both forecasting and hotspot prediction, the
``fake'' indicator models (with random noise as auxiliary features) offered
essentially no improvement over the pure AR model, which provides a strong
rejection (informally speaking) of the implicit regularization hypothesis.

\begin{figure}[t]
\includegraphics[width=\columnwidth]{fig/hot-1.pdf}
\caption{Area under the curve for each forecast model averaged over all
  prediction dates and HRRS.}
\label{fig:hotspot}
\end{figure}

\subsection{Performance in Up, Down, Flat Periods}

\begin{figure}[t]
  \includegraphics[width=\columnwidth]{fig/upswing-histogram-1.pdf}
  \caption{Histogram of the difference in WIS for the AR model relative to each
    forecaster. Larger values are better for the indicator assisted model.}
  \label{fig:up-down-flat}
\end{figure}

@@

@@

Can auxiliary indicators improve COVID-19 forecasting and hotspot prediction
models?  Our answer, based on analyzing five auxiliary indicators from the  
COVIDcast API (defined using from medical insurance claims, internet-based
surveys, and Google search trends) is undoubtedly ``yes''. However, there are 
levels of nuance to such an answer that must be explained.  None of the
indicators that we have investigated appear to be the ``silver bullet'' that 
one might have hoped for, revolutionizing the tractability of the prediction 
problem, rendering it easy when it was once hard (in the absence of auxiliary 
information).  Rather, the gains in accuracy from the indicator models (over an
autoregressive model based only on past case rates) appear to be nontrivial, and
consistent across modes of analysis, but modest.  In forecasting, the indicator 
models appear to be most useful in periods in which case rates are flat or
trending down, rather than periods in which case rates are trending up (as one 
might hope to see is the benefit provided by a hypothetical ``leading
indicator'').

>>>>

\begin{figure}[t]
\includegraphics[width=\columnwidth]{fig/trajectory.pdf}
\caption{Forecast from an autoregressive model for the New York City HRR made 
  on October 15 (vertical line).  The fan displays 50\%, 80\% and 95\% intervals
  while the orange curve shows the median forecast. The black curve represents
  ``finalized'' data, as reported in May 2021.} 
\label{fig:trajectory}
\end{figure}

@@

In addition to this pure autoregressive model, we also consider five
logistic models of the form:  
\begin{multline}
\label{eq:hotspot_ar_x}
\logit \big( \P(Z_{\ell,t+a} = 1 \,|\, Y_{\ell,s}, X_{\ell,s}, s \leq t) \big)
= \\ \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y^\Delta_{\ell,t-7j} +  
\sum_{j=0}^2 \gamma^{a,\tau}_j X^\Delta_{\ell,t-7j},
\end{multline}
where we use \smash{$X^\Delta_{\ell,t} = (X_{\ell,t} - X_{\ell,
    t-7})/(X_{\ell,t-7})$}, and again $X_{\ell,t}$ stands for any of the five  
auxiliary indicators at location $\ell$ and time $t$.  We fit the above models,
\eqref{eq:hotspot_ar}, \eqref{eq:hotspot_ar_x}, using logistic regression,   
pooling all 306 HRRs and using a 21-day trailing window for the training data.

@@

WIS is typically defined for quantile-based forecasts where the quantile levels
are symmetric around 0.5.  This is the case for our choice in
\eqref{eq:quantile_levels}.  Let $F$ be a forecaster comprised of predicted
quantiles $q_\tau$ parametrized by a quantile level $\tau$.  In the case of
symmetric quantile levels, this is equivalent to a collection of central
prediction intervals $(\ell_\alpha, u_\alpha)$, parametrized by an exclusion
probability $\alpha$. The WIS of the forecaster $F$, evaluated at the target
variable $Y$, is defined by:
\begin{equation}
\label{eq:wis_intervals}
\WIS(F,Y) = \sum_\alpha \Big\{\alpha(u_\alpha - \ell_\alpha) + 2 \cdot
\dist(Y, [\ell_\alpha, u_\alpha])\Big\},  
\end{equation}
where $\dist(a, S)$ is the distance between a point $a$ and set $S$ (the
smallest distance between $a$ and an element of $S$).  Note that, corresponding 
to \eqref{eq:quantile_levels}, the exclusion probabilities are $\alpha \in
\{0.05, 0.2, 0.5, 1\}$, resulting in 4 terms in the above sum.  By 
straightforward algebra, it is not hard to see WIS has an alternative 
representation in terms of the predicted quantiles themselves:
\begin{equation}
\label{eq:wis_quantiles}
\WIS(F,Y) = 2 \sum_\tau \phi_\tau(Y - q_\tau), 
\end{equation}
where $\phi_\tau(x) = \tau |x|$ for $x \geq 0$ and $\phi_\tau(x) = (1-\tau)
|x|$ for $x<0$, which is often called the ``tilted absolute'' loss.  While
\eqref{eq:wis_quantiles} is more general (it can accomodate asymmetric quantile
levels), the first form in \eqref{eq:wis_intervals} is typically preferred in
presentation, because the score nicely decouples into a ``sharpness'' component
(first term in each summand) and ``under/overprediction'' component (second term
in each summand).  The second form given in \eqref{eq:wis_quantiles} is
especially noteworthy in our current study as it reveals WIS is the same as the
quantile regression loss that we use to train our forecasting models (i.e., our
models are fit to optimize WIS averaged over the training set).

@@

We begin with a summary of the high-level conclusions.  Here, and in the
following subsections, we will use ``AR'' to refer to the pure autoregressive
model both in forecasting, \eqref{eq:forecast_ar}, and in hotspot prediction,
\eqref{eq:hotspot_ar} (the reference to the prediction task should always be
clear from the context). We will also use the name of an auxiliary
indicator---namely ``CHNG-CLI'',  ``CHNG-COVID'', ``CTIS-CLI-in-community'',
``DV-CLI'', or ``Google-AA''---interchangeably with the model in forecasting, 
\eqref{eq:forecast_ar_x}, or hotspot prediction, \eqref{eq:hotspot_ar_x}, that
uses this particular indicator as a feature (the meaning should be clear from
the context).  So, for example, the CHNG-CLI model in forecasting is the one in  
\eqref{eq:forecast_ar_x} that sets $X_{\ell,t}$ to be the value of the CHNG-CLI 
indicator at location $\ell$ and time $t$.
% fits the coefficients based on the training data, and makes its predictions by
% setting $X_{\ell,t}$ to the appropriate CHNG-CLI values.  
Finally, we use the term ``indicator model'' to refer to any one of the ten
models of the form \eqref{eq:forecast_ar_x}, \eqref{eq:hotspot_ar_x} (five from
each of the forecasting and hotspot prediction tasks).

@@

Figure~\ref{fig:forecast} displays the results for forecasting, stratified by
ahead value and averaged over all HRRs and forecast dates (June 9--December 31,
2020).  Shown is the average WIS for each forecast model divided by the averaged
WIS from a baseline model, which is basically a flat-line forecaster (its
median forecast for $Y_{\ell,t+a}$ is always $Y_{\ell,t}$, with predicted
quantiles defined around this flat line based on historical variation). This is
the same baseline model as that used in the COVID-19 Forecast Hub.  Here, we  
divide by the average WIS of the baseline model in order to put the y-axis on an 
interpretable scale.  We can see that all curves are below 1, which means
(smaller WIS is better) that all of the models, including AR, outperform the
baseline on average over the forecasting period.  On the other hand, the models 
deliver at best an improvement of about 20\% in mean WIS over the baseline
model, with this gap narrowing to only 10\% at the largest ahead values,
illustrating the difficulty of the forecasting problem.

@@

\begin{figure}[t]
  \includegraphics[width=\columnwidth]{fig/fcast-1.pdf}
  \caption{Average WIS for each forecast model, over all forecast dates and all
    HRRs, divided by the average WIS achieved by a baseline model (a
    probabilistic version of the flat-line forecaster).  Here and in all figures
  we abbreviate CTIS-CLI-in-community by CTIS-CLIIC.}  
  \label{fig:forecast}
\end{figure}

Figure~\ref{fig:hotspot} displays the results for hotspot prediction, again
stratified by ahead value and averaged over all HRRs and prediction dates (June 
16--December 31, 2020).  We can see many similarities to the forecasting
results (now larger AUC is better); for example, CHNG-COVID and DV-CLI offer the
biggest improvement over AR, and all models including AR degrade in performance 
towards the baseline (this is now just a classifier based on random guessing,
which achieves an AUC of 0.5) as the ahead values grow.  A clear difference,
however, is that the CHNG-CLI model performs quite well in hotspot prediction,
close to the best-performing indicator models for many of the ahead values. This
may be because volatility in the CHNG-CLI indicator plays less of a role in the
associated logistic model's predicted probabilities (in general, a sigmoid
function can absorb a lot of the variability in its input).

@@

To test this hypothesis, we reran all of the prediction experiments but with
$X_{\ell,t}$ in the each indicator models in \eqref{eq:forecast_ar_x},
\eqref{eq:hotspot_ar_x} replaced with suitable random noise (bootstrap samples
from a signal's historical distribution).  The results, shown and explained more
precisely in the supplement, are vastly different (worse) than the original set
of results.  In both forecasting and hotspot prediction, the ``fake'' indicator
models (with random noise as auxiliary features) offered essentially no
improvement over the pure AR model, which provides a strong rejection
(informally speaking) of the implicit regularization hypothesis.

Along the lines of regularization, it is also worth noting that the use of
$\ell_1$ regularization (tuned by cross-validation) in fitting any of the models
in \eqref{eq:forecast_ar}, \eqref{eq:forecast_ar_x}, \eqref{eq:hotspot_ar}, 
\eqref{eq:hotspot_ar_x} did not improve their generally performance at all
(experiments not shown).  This is likely due to the fact that the number of
training samples is quite large compared to the number of features (6,426
training samples and only 3--6 features).

\begin{figure}[t]
\includegraphics[width=\columnwidth]{fig/hot-1.pdf}
\caption{Area under the curve for each hotspot prediction model, calculated over
  all prediction dates and all HRRs.} 
\label{fig:hotspot}
\end{figure}

\subsection{Performance in Up, Down, and Flat Periods}

The course of the pandemic has played out quite differently across space and
time. Aggregating case rates nationally shows three pronounced waves, but the
behavior is more nuanced at the HRR level.  Figure \ref{fig:trajectory} is a
single example of a forecast in a period of relatively flat case trends, as New
York City enters what would become its second wave. The AR forecaster's 50\%
prediction interval contains this upswing, but its forecasted median is clearly
below the finalized case data.  Unfortunately, this behavior is fairly typical
of all forecasters: during upswings, the forecasted median tends to fall below 
the target, while the reverse is true during downswings.

Figure~\ref{fig:up_down_flat} shows histograms of the differences in WIS of the 
AR model and each indicator model, where we stratify these differences by
whether the target occurs during a period of increasing cases rates (up),
decreasing case rates (down), or flat case rates (flat). To define the
increasing period, we use the same definition we used for the hotspot task in
Table \ref{tab:analysis_tasks}.  Therefore all hotspots are ``up'', while all
non-hotspots are either ``flat'' or ``down''.  For the ``down'' scenario, we 
simply use the opposite of the hotspot definition: $Y_{\ell,t}$ decreases by
more than 25\% relative to the preceding week.

While the performance of all forecasters, including AR, will generally degrade
in up periods, different models exhibit different and interesting patterns.
CHNG-CLI, CHNG-COVID, Google-AA, and especially CTIS-CLI-in-community have
large right tails (displaying improvements over AR) over the down periods.
Google-AA and CTIS-CLI-in-community have large right tails during the flat
periods. CHNG-CLI and DV-CLI have large left tails (poor forecasts relative to
AR) in flat and up periods.  Google-AA is the only model that outperforms
the AR model, on average, over up periods.  Overall, the indicators seem to help
more during flat or down periods than up periods, with the exception of
Google-AA.

\begin{figure}[t]
  \includegraphics[width=\columnwidth]{fig/upswing-histogram-1.pdf}
  \caption{Histogram of the difference in WIS for the AR model and that for each
    indicator model, stratified by up, down, or flat periods, measured in terms
    of case activity. (Larger values are better for the indicator model.) The
    y-axis is on the log scale to emphasize tail behavior.} 
  \label{fig:up_down_flat}
\end{figure}

The supplement pursues this analysis further.  For example, we examine
classification accuracy and log-likelihood for the hotspot task, and find a
similar phenomenon: the indicators considerably improve accuracy or
log-likelihood during flat or down periods, with more mixed behavior during up
periods, with the CHNG-CLI, CHNG-COVID, and DV-CLI leading to decreased
performance in particular.

% Essentially, during downswings, improved performance is strongly correlated
% with predicing larger declines than the AR model. The AR model tends to
% overpredict relative to the decline, but adding indicators tends to diminish
% this effect. During upswings, the AR model tends to underpredict, but when the
% addition of an indicator improves performance (a more rare occurrence) it is 
% rarely due to higher median predictions.

% RJT: Commented out the last paragraph for now, because this is kind of a tough
% thing to explain succintly. Can revisit if we really want to include it here,
% or just let them find it in the supplement?

@@

As described in the methods section, each of the indicators we examine could be
said to measure aspects on disease progression that would precede a positive
test. That is, we imagine that these signals should ``lead'' cases. It is
entirely reasonable to imagine that, prior to an increase of confirmed COVID-19
tests reported by public health in a particular location, we would see an
increase in insurance claims for COVID-19-related medical visits.  However, it
may well be the case that this leading behavior is different during different
periods. In fact, we find that the ``leadingness'' of an indicator (degree to
which it leads) tends to be more pronounced during down and flat periods than up
periods, a plausible explanation for the decreased performance noted above.

In order to measure the leadingness on af indicator, we examine its cross
correlations to case rates over a short time window

 the cross
correlation function. Essentially this examines the Pearson correlation between

shifted version of the indicator and the case rate. So if $X_{\ell,t}$ is
shifted forward by 7 days, and the magnitude of the correlation of that shifted series with
$Y_{\ell,t}$ (unshifted) is high, we would say that $X_{\ell}$ leads $Y_{\ell}$
by 7 days. This process is repeated for a range of shifts, forward and backward,
resulting in leadingness and laggingness scores. Further details are given in
the supplement.

Figure \ref{fig:leading-lagging} shows the correlations between the difference
in the WIS of the AR model relative to $F$ with the leading and laggingness
scores stratified by whether the target is classified as up, down, or flat. One
would expect that increases in $\WIS(AR) - \WIS(AR)$ would be positively
correlated with leadingness. This relationship is strongest during down periods
and weakest during up periods with flat in the middle. In fact, for each
indicator, the strength of correlations between improved performance and
leadingness can be ordered down $>$ flat $>$ up. The relationship with
laggingness is the opposite, though less pronounced. CHNG-CLI's forecast
performance doesn't correlate very strongly with laggingness while any
laggingness in Google-AA during down and flat periods results in realtively
worse forecasting performance.

Because a time series can, to a degree, both simultaneously leading and lagging,
we also examine the correlation between performance and the difference between
leadingness and laggingness. Figure \ref{fig:diff-btw-leading-lagging} shows
this behavior. Again, indicators that are more leading then lagging tend to
are positively correlated with improved relative performance. This relationship
is most pronounced during periods of decreasing case rates and attenuates in
other scenarios.

@@

Can auxiliary indicators improve COVID-19 forecasting and hotspot prediction
models?  Our answer, based on analyzing five auxiliary indicators from the  
COVIDcast API (defined using from medical insurance claims, internet-based
surveys, and Internet search trends) is undoubtedly ``yes''. However, there are  
levels of nuance to such an answer that must be explained.  None of the
indicators that we have investigated appear to be the ``silver bullet'' that 
one might have hoped for, revolutionizing the tractability of the prediction 
problem, rendering it easy when it was once hard (in the absence of auxiliary 
information).  Rather, the gains in accuracy from the indicator models (over an
autoregressive model based only on past case rates) appear to be nontrivial, and
consistent across modes of analysis, but modest.  In forecasting, the indicator 
models appear to be most useful in periods in which case rates are flat or
trending down, rather than periods in which case rates are trending up (as one 
might hope to see is the benefit provided by a hypothetical ``leading
indicator'').
